Optimal updating factor in Wang-Landau and metadynamics simulations

### 2016-04-19 doc/invt.tex
add massq.gp

@@

>>>>

\subsubsection{\label{sec:mass_distr}
Characterization of the optimal schedule}

We can characterize the optimal schedule
geometrically
through the normalized mass distribution
along $q$ as
\begin{equation}
  m(q)
  =
  \frac{
    \sqrt{
      \sum_{k = 2}^n \Gamma_k \, \lambda_k^2 \, u_k^2(q)
    }
  }
  {
    \int_0^{ q(T) }
      \sqrt{
        \sum_{k = 2}^n \Gamma_k \, \lambda_k^2 \, u_k^2(q')
      }
      \, d q'
  }
  ,
\label{eq:mass_distr}
\end{equation}
%
such that
$\int_0^{q(T)} m(q) = 1$.
%
As shown in Fig. ,
the distribution peaks at $q = q(T)$,
and decreases with the difference $q(T) - q$.
%
According to Eq. \eqref{eq:q_opt},
the ordinate of the distribution curve at $q$
gives $1/(T\alpha) = 1/(T \dot q)$,
and the area under the curve in the region $q' < q$
gives $t/T$.
%
From this picture, we can draw several conclusions.

\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=0.9\linewidth]{fig/massq.pdf}
  }
  \caption{
    \label{fig:massq}
    Geometric characterization of the optimal schedule,
    $\alpha(t)$,
    via the normalized mass distribution, $m(q)$.
    %
    \note{The figure was produced by \texttt{doc/fig/massq.gp}.
    }%
  }
\end{center}
\end{figure}

First,
the optimal schedule has the following scaling property.
%
For two simulations of lengths $T_1$ and $T_2$,
if $q(T_1) = q(T_2)$
and they share the same $\lambda_k$'s and $\Gamma_k$'s,
the updating magnitudes at the corresponding times,
$t_1/T_1 = t_2/T_2$,
can be related as $T_1 \, \alpha_1 = T_2 \, \alpha_2$.
%
In other words,
if $\alpha(t_1)$ is
an optimal schedule of the simulation of length $T_1$,
then
$\frac{T_1}{T_2} \alpha\left( \frac{ T_1 } { T_2 } t_2 \right)$
is an optimal schedule of the simulation of length $T_2$.
%
The asymptotic errors are related by
$T_1 \, E_{A1} = T_2 \, E_{A2}$.

@@

\subsection{\label{sec:deplength}
Dependence on the simulation length}

The parameter, $\lambda$, is equal to
the difference
$$
  \lambda
  =
  \Delta \left( \frac{ 1 } { T \, \alpha } \right)
  \equiv
  \left.
  \frac { 1 } { T \, \alpha(t) }
  \right|_{t = 0}^{t = T}
  .
$$
which corresponds to the height difference
of the normalized mass distribution
$
\left. m(q) \right|_{q = 0}^{q = q(T)}
$
in the optimal schedule
(cf. Sec. \ref{sec:mass_distr}).
%
Therefore, we expect the optimal $\lambda$
to decrease with the span $q(T)$,
hence the simulation length $T$,
if the distribution has a long tail.
